A novel route to the spontaneous formation of porous crystals via viscoelastic phase separation

### 2016-03-22 nature/response.tex

\begin{referee}
2) Page 5, second paragraph: the text states that the vol frac of 68\% corresponds to the vol frac of the stable crystal phase. This contradicts the phase diagram shown in Fig 1A, which has the coexisting crystal vol frac at about 0.72 or so - far over toward the right-hand axis. This should be quite distinct from 0.68. There may be good reasons for the difference between the measured 0.68 and the expected equilibrium value, but this should be discussed. 
\end{referee}

The Referee here points a real inconsistency in our previous analysis. We tracked our mistake down to our estimate of the volume fraction stemming from the poor precision on the colloid diameter obtained either from diffusion constant measurement ($2.3\pm0.5\,\mu$m) or from the position of the first peak of $g(r)$ ($2.2\pm0.5\,\mu$m). In the present version, we instead matched the stable crystal phase volume fraction to the asymptotic value of the local volume fraction of 12-neighboured particles in the crystallising sample. This yields a much more precise value for the colloid diameter ($2.25\,\mu$m). Our method is detailed in Supplementary Information.

To cross-check the validity of this method, we extended both our experimental measurements and theoretical predictions to be able to compare the phase boundaries in all samples. On the theory side, we computed the position of the liquid side of both binodal and spinodal lines. For this, we had use a hard sphere equation of state that describes well the approach to jamming. On the experimental side, we used a Gaussian approximation of the density field to measure the amount of gas and liquid and deduce the volume fraction in the liquid phase. The updated Fig 1a shows a good agreement between the liquid-gas tie lines measured experimentally and the predicted binodal line. \textcolor{blue}{Here we note that the volume fraction of the dense phase is an average over a certain range of the number of neighbours. As shown in Fig. 1d. 12 neighbours is a maximum in compaction and thus the dense phase can only have lower volume fraction.}

### 2016-03-23 crystal-gel/nature/response.tex

\begin{referee}
2) Page 5, second paragraph: the text states that the vol frac of 68\% corresponds to the vol frac of the stable crystal phase. This contradicts the phase diagram shown in Fig 1A, which has the coexisting crystal vol frac at about 0.72 or so - far over toward the right-hand axis. This should be quite distinct from 0.68. There may be good reasons for the difference between the measured 0.68 and the expected equilibrium value, but this should be discussed. 
\end{referee}

The Referee here points a real inconsistency in our previous analysis. We tracked our mistake down to our estimate of the volume fraction stemming from the poor precision on the colloid diameter obtained either from diffusion constant measurement ($2.3\pm0.5\,\mu$m) or from the position of the first peak of $g(r)$ ($2.2\pm0.5\,\mu$m). In the present version, we instead matched the stable crystal phase volume fraction to the asymptotic value of the local volume fraction of 12-neighboured particles in the crystallising sample. This yields a much more precise value for the colloid diameter ($2.25\,\mu$m). Our method is detailed in Supplementary Information.

To cross-check the validity of this method, we extended both our experimental measurements and theoretical predictions to be able to compare the phase boundaries in all samples. On the theory side, we computed the position of the liquid side of both binodal and spinodal lines. For this, we had use a hard sphere equation of state that describes well the approach to jamming. On the experimental side, we used a Gaussian approximation of the density field to measure the amount of gas and liquid and deduce the volume fraction in the liquid phase. The updated Fig 1a shows a good agreement between the liquid-gas tie lines measured experimentally and the predicted binodal line. \textcolor{blue}{Here we note that the volume fraction of the dense phase is an average over a certain range of the number of neighbours. As shown in Fig. 1d. 12 neighbours is a maximum in compaction and thus the dense phase can only have lower volume fraction.}

@@

To address the concern about the reproducibility of our kinetic pathway, we decided to make additional independent experiments
by using a sample with a similar composition, which is inserted into an ordinary glass capillary tube. We thus confirm the reproducibility of this type of crystallization pathway. An example of crystal-gel porous  network obtained through our new capillary experiments is found in new Supplementary Fig. S5. All details about the experiments are also included the Supplementary Information. We would like to thank the Referee for stimulating this additional investigation.

\begin{referee}
2) From the SI movie, I don't recognize the final reported structure to be a crystal gel. These final structures seem instead to be crystallites of finite size at equilibrium. The crystallites diffusive minimally because they are jammed against their neighbors. The results seem similar to Anderson and Lekkerkerker's (ref 20) explanation of Ostwald ripening. The changes in crystal size at late times, when the liquid to gas to crystal mechanism is proposed, appear that they could be equally well described as a process in which larger crystallites grow at the expense of smaller ones, through the gas phase.
\end{referee}

The impression of the reviewer that the final structures are made of crystallites of finite size may come from the fact that the percolated network structure cannot be seen by 2D sectioning. The Movies show only a thin slice of the system which corresponds to the particle that are in focus. 
As can be seen in Fig. 2b (left panel), the crystal gel has a bicontinuous porous structure. 
To show the connectivity of the network clearly with a particle-level resolution, we have also made Supplementary Figure S5, highlighting the crystal network. 
We also include a movie of a thin slice with a finite thickness (Supplementary Movie 3).
We hope that this will clarify the situation and convince the reviewer of the fact that our sample indeed forms a porous network of crystals.

### 2016-03-23 nature/Supplementary.tex
New figures.

The use of a reservoir of a salt solution through a semi-permeable membrane, allowed us to follow the entire microscopic dynamics in absence of spurious flows, from the first stages of phase separation to the formation of the crystal-gel state. In order to confirm the reproducibility of the crystal-gel state, we conducted experiments in a more traditional capillary setting, where all components of the suspension are pre-mixed in a sealed cell. A typical example is
shown in Fig.~\ref{fig:capillary}, where a system with the colloid volume fraction of $\approx 0.33$ and the polymer concentration of $0.40\,$mg/g is considered.
The figure shows the time evolution of the sample at three different times, just after phase separation ($t=340\,\tau_{\rm B}$), and during
crystal-gel formation ($t=2600\,\tau_{\rm B}$ and $t=6650\,\tau_{\rm B}$). The top-row shows images from quasi two-dimensional confocal slices, while the bottom-row shows
the three-dimensional reconstruction after single-particle tracking, where crystalline particles are highlighted in both size (large) and color (red).

@@

>>>>

The use of a reservoir of a salt solution through a semi-permeable membrane, allowed us to follow the entire microscopic dynamics in absence of spurious flows, from the first stages of phase separation to the formation of the crystal-gel state. In order to confirm the reproducibility of the crystal-gel state, we conducted experiments in a more traditional capillary setting, where all components of the suspension are pre-mixed in a sealed cell. A typical example is
shown in Fig.~\ref{fig:capillary}, where a system with the colloid volume fraction of $\approx 0.33$ and the polymer concentration of $0.40\,$mg/g is considered.
The figure shows the time evolution of the sample at three different times, just after phase separation ($t=340\,\tau_{\rm B}$), and during
crystal-gel formation ($t=2600\,\tau_{\rm B}$ and $t=6650\,\tau_{\rm B}$). The top-row shows images from quasi two-dimensional confocal slices, while the bottom-row shows
the three-dimensional reconstruction after single-particle tracking, where crystalline particles are highlighted in both size (large) and color (red), surface particles
are depicted in blue, gas in green and liquid in white.

@@

\section*{Supplementary Movie 3}
Same as Supplementary Movie 2, but with computer rendered trajectories obtained from the tracking of the colloidal particles.
Particles are colored according to the following code: crystal (red), surface (blue), gas (green) and liquid (white).
For display purposes, crystalline particles and gas particles are respectively 400\% and 50\% bigger than other particles (liquid or surface).
